---
title: "Data Exploration"
output: pdf_document
date: "2025-10-08"
author: "Team 10"
---

## Introduction

This report provides a short exploratory data analysis of the IMDb movie dataset.  
The purpose is to better understand the structure of the data, identify key variables, check for missing values, and visualize general patterns such as rating distributions and trends over time.

---

## Loading the Data

The first step is to load the required R packages and import the dataset.  
Make sure the file path to your dataset is correct before running the code.

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)

setwd("../")
imdb <- read_csv("gen/temp/imdb_movies.csv")
head(imdb)
```
## Dataset Overview

Before analyzing the data, we first check how many observations (rows) and variables (columns) the dataset contains.
Then, we create a summary table showing the variable names, data types, missing values, example entries, and short descriptions.
```{r message=FALSE, warning=FALSE, echo=FALSE}
descriptions <- c(
  "Movie ID",
  "Duration of the movie in minutes",
  "Genre of the movies",
  "Release year",
  "Average IMDB rating",
  "Number of votes",
  "Director of the movie"
)

summary_df <- data.frame(
  Variable = names(imdb),
  Type = sapply(imdb, class),
  Missing_Data = sapply(imdb, function(x) sum(is.na(x))),
  Example = sapply(imdb, function(x) paste(head(x, 1), collapse = ", ")),
  Description = descriptions[1:length(names(imdb))],
  row.names = NULL
)

summary_df %>%
  kable(caption = "Overview of Variables in IMDb Dataset") %>%
  kable_styling(full_width = FALSE, position = "left")
```

## Dimensions of the Dataset
```{r message=FALSE, warning=FALSE, echo=FALSE}
variable_1 <- nrow(imdb)
variable_2 <- ncol(imdb)

```
We have `r variable_1` observations and we have `r variable_2` columns.

## Summary Statistics

Here we calculate basic descriptive statistics, including the minimum, quartiles, median, mean, maximum, and number of missing values, for selected numeric variables.
```{r message=FALSE, warning=FALSE, echo=FALSE}
summary_table <- imdb %>%
  summarise(across(
    c(runtimeMinutes, startYear, averageRating, numVotes),
    list(
      Min = ~min(.x, na.rm = TRUE),
      `1st Qu.` = ~quantile(.x, 0.25, na.rm = TRUE),
      Median = ~median(.x, na.rm = TRUE),
      Mean = ~mean(.x, na.rm = TRUE),
      `3rd Qu.` = ~quantile(.x, 0.75, na.rm = TRUE),
      Max = ~max(.x, na.rm = TRUE),
      Missing = ~sum(is.na(.x))
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  tidyr::pivot_longer(everything(),
                      names_to = c("Variable", ".value"),
                      names_sep = "_")

knitr::kable(summary_table, caption = "Summary Statistics for IMDB Variables", digits = 1) %>%
  kable_styling(full_width = FALSE, position = "left")
```

## Visualizations

In this section, we visualize the main characteristics of the IMDb dataset.
We focus on the distribution of ratings, trends over time, and relationships between variables.
```{r message=FALSE, warning=FALSE, echo=FALSE}
imdb_complete <- imdb %>%
  filter(!is.na(averageRating), !is.na(startYear),
         !is.na(runtimeMinutes), !is.na(numVotes))
```

### Distribution of Average Movie Ratings

This histogram shows how movie ratings are distributed across the dataset.
```{r fig.align='center', message=FALSE, warning=FALSE, echo=FALSE}
ggplot(imdb_complete, aes(x = averageRating)) +
  geom_histogram(bins = 10, fill = "#F5C518", color = "black") +
  labs(
    title = "Distribution of Average Movie Ratings",
    x = "Average Rating",
    y = "Number of Movies"
  ) +
  scale_x_continuous(breaks = seq(from = min(imdb_complete$averageRating, na.rm = TRUE),
                                 to = max(imdb_complete$averageRating, na.rm = TRUE),
                                 by = 1)) +
  theme_minimal()
```
### Average Rating by Year

The following line chart shows how the average rating has evolved over time.
This can reveal general trends or shifts in audience perception.
```{r fig.align='center', message=FALSE, warning=FALSE, echo=FALSE}
# Average rating per year
imdb_yearly <- imdb_complete %>%
  group_by(startYear) %>%
  summarise(mean_rating = mean(averageRating, na.rm = TRUE))

ggplot(imdb_yearly, aes(x = startYear, y = mean_rating)) +
  geom_line(color = "#F5C518") +
  labs(
    title = "Average Movie Rating by Year",
    x = "Release Year",
    y = "Average Rating"
  ) +
  scale_x_continuous(breaks = seq(from = min(imdb_yearly$startYear, na.rm = TRUE),
                                 to = max(imdb_yearly$startYear, na.rm = TRUE),
                                 by = 10)) +
  theme_minimal()
```
### Relationship Between Number of Votes and Ratings

Here we visualize the relationship between the number of user votes and the average rating.
Highly popular movies might show different rating behavior compared to less known films.
```{r fig.align='center', message=FALSE, warning=FALSE, echo=FALSE}
# Relation between votes and rating
ggplot(imdb_complete, aes(x = numVotes, y = averageRating)) +
  geom_point(alpha = 0.2, color = "#F5C518") +
  labs(
    title = "Relationship between Number of Votes and Average Rating",
    x = "Number of Votes",
    y = "Average Rating"
  ) +
theme_minimal()
```
### Distribution of Movie Runtimes

This histogram focuses on typical movie lengths (0–300 minutes), giving insight into runtime trends.
```{r fig.align='center', message=FALSE, warning=FALSE, echo=FALSE}
# Filter between 0 - 300 minutes
imdb_runtime <- imdb_complete %>%
  filter(runtimeMinutes > 0, runtimeMinutes <= 300)

# Histogram of length of movies
ggplot(imdb_runtime, aes(x = runtimeMinutes)) +
  geom_histogram(bins = 30, fill = "#F5C518", color = "black") +
  labs(
    title = "Distribution of Movie Runtimes (0-300 minutes)",
    x = "Runtime (minutes)",
    y = "Number of Movies"
  ) +
  theme_minimal()
```

## Conclusion

From this exploratory analysis, we can conclude the following:

- The IMDb dataset contains information on several million movies, covering multiple attributes such as ratings, release year, runtime, and popularity.
- Most movies have ratings clustered around a central value (often between 6 and 7).
- Average ratings appear relatively stable across years, though minor trends may exist. In the first period, before 1914, it isn't stable, probably because of a small amount of movies.
- Movies with more votes tend to be those that are more popular or widely known.
- Typical movie runtimes lie between 80 and 130 minutes, with only a few outliers.

This dataset provides a valuable foundation for further analysis, such as predicting director success on a specific directors experience, and exploring genre effects.
