---
title: "document_project_data_preparation_group10_PDFmarkup."
author: "Group 10"
date: "2025-09-04"
output:
word_document: default
pdf_document: default
---
Daan Bankers              snr:2100281
Luc Vesters				        snr:2145531
Seppe Bogers				      snr:2080522
Melvin Hong				        snr:2146253
Wouter van der Linden			snr:2116790
Marvin  Herwig			    	snr:2138521


```{r analysis_packages, echo = FALSE, message=FALSE, warning=FALSE}
#Loading packages
library(dplyr)
library(GGally)
library(ggplot2)
library(tidyr)
library()
```

# Does directional experience influence IMDb ratings? 
## Examining how director experience predicts IMDb ratings and how this relationship differs between drama and action movies.

### Introduction

#### Research Question

To what extent does a director’s experience( measured as the cumulative number of minutes of all movies previously directed) predict a movie’s IMDb rating, and is this relationship moderated by movie genre (drama vs. action)?

#### Academically Relevant

This study contributes to research on the movie industry by examining how directors’ cumulative experience predicts movie evaluations. We use total directed minutes, which captures experience more precisely. Then the length of the movies give less experience when it’s rather short than a few very long ones. The IMDb ratings capture consumer evaluation patterns. Furthermore, genre could play a crucial role, given that the genre drama is more critically evaluated than the genre action, which is more popular. This suggests genre could moderate this relationship.

#### Managerially Relevant

For studios and investors facing uncertainty, insights into how directors experience influences audience rating are useful. If experience predicts higher quality, the decision for an experienced director could be made easier. Recognizing genre differences helps managers allocate talent effectively. While technical production could dominate the genre drama. This reduces risk and aligns creative choices with market demands.

#### Statistical Testing Method

Multiple regression analysis is suitable for testing a metric independent variable(director experience) and a metric independent variable(IMDb rating), while including a nonmetric moderator (genre: drama vs. action). By applying this method, this estimates the direct and indirect effect in a single model. It also allows controlling for other factors, making the results more interpretable.



### Creating  the raw data set

#### packages and folders
The packages that this projects needs to run are listed in the installpackages.R document in the rootdirectory of this project. The download.R document makes also sure that right working directorys will be set and that the folders and subfolders are created and put in the right place.


#### datasets that will be used in this project

This project uses datasets from IMDB movies. The following datasets from IMDB movies are used in this project:
-name.basics.tsv
-title.basics.tsv
-title.crew.tsv
-title.ratings.tsv

#### adjustments to create the raw dataset

The title.basics.tsv file will be filtered immediately after it has been downloaded. This is because the dataset contains a variable called "primaryTitle" and it describes the type of the video material. That is the reason why "movie" is filtered from the primaryTitle variable, because tv series, shorts and other types of video material should not be included in this analysis. Only the columns that were relevant for our research question (tconst, runtimeMinutes, genres, startYear) were taken from the title.basics.tsv. This because of the size of the title.basics.tsv file. After filtering the title.basics.tsv dataset, the "name.basics.tsv", the "title.crew.tsv", the "title.ratings.tsv" and the filtered version of "title.basics.tsv" were saved as ".csv" files/datasets. These files/datasets are merged using a full join because the maximum amount of data well be retained this way. Finally, this big merged datset will be saved as "imdb_movies_dataset". (The way of merging and saving the imdb_movies_dataset is shown in the clean_and_merge.R rscript)     
`
#### operationalisation tabels 
Operationalisation tables are made to provide more information about the variables that are included in the dataset. The operationalisation tables are called: name_basics_cols, title_crew_cols, title_basics_cols and title_ratings_cols( these tables are shown below). 

```{r preparing_data, include=TRUE}


# Column description
name_basics_cols <- tibble(
  column = c("nconst", "primaryName", "birthYear", "deathYear", "primaryProfession", "knownForTitles"),
  description = c(
    "alphanumeric unique identifier of the name/person",
    "name by which the person is most often credited",
    "in YYYY format",
    "in YYYY format if applicable, else '\\N'",
    "the top-3 professions of the person",
    "titles the person is known for"
  )
)

title_crew_cols <- tibble(
  column = c("tconst", "directors", "writers"),
  description = c(
    "alphanumeric unique identifier of the title",
    "director(s) of the given title",
    "writer(s) of the given title"
  )
)

title_basics_cols <- tibble(
  column = c("tconst", "titleType", "primaryTitle", "originalTitle", "isAdult", "startYear", "endYear", "runtimeMinutes", "genres"),
  description = c(
    "alphanumeric unique identifier of the title",
    "the type/format of the title (e.g. movie, short, tvseries, tvepisode, video, etc)",
    "the more popular title / the title used by the filmmakers on promotional materials at the point of release",
    "the original title, in the original language",
    "0: non-adult title; 1: adult title",
    "represents the release year of a title. In the case of TV Series, it is the series start year",
    "TV Series end year. '\\N' for all other title types",
    "primary runtime of the title, in minutes",
    "includes up to three genres associated with the title"
  )
)

title_ratings_cols <- tibble(
  column = c("tconst", "averageRating", "numVotes"),
  description = c(
    "alphanumeric unique identifier of the title",
    "weighted average of all the individual user rating",
    "number of votes the title has received"
  )
)


```

#### handling missing values and class problems

The missing values in the "imdb_movies_dataset" were indicated with the as "\\N". the code used indicated the missing values with NA. This because  functions like "rm.na" or "is.na"  were easier to work with and it makes analysis easier. Finally the class of runtimeMinutes changed into an integer and the class of startYear changed to numeric. Both these variables were classified as characters. by changing them, stats could be performed and proper comparisions could be performed as well. These changes were saved in a new csv file called: "imdb_movies.csv" 

### cleaning data for the direct effect
From the "imdb_movies.csv" dataset, the 5% extreme values in the dataset will be ignored. This because the results are not normally divided in the when we do not ignore the 5% extreme values. The dataset contains some extreme points with a totalruntime over the 15000 minutes. Normality is needed when we want to run a muliple linear regression. That is the reason why the 5% extreme values will be ignored. This project only uses films that are created from 1990 onwards. The digital revolution started in 1990 and changed the film industry as well. New digital visual effects could be used in films because of this revolution. The observations are also filtered in a way that the na's are ignored as well. A seperate row function makes sure that every director has its own row, because there are observations that included multiple directors. Finally, three columns are added to the dataset. The column "total_runtime" which is the sum of the runtime in minutes. This way we were able to get the total_runtime of film that a director produced in its career. The second column is "avg_rating" which is the mean of the average_rating of all films that the director created. The last column is n_films column which counts the amount of films that a director has made. All these adjustements are saved in a .csv file called "imdb_movies_direct.csv".


#hier mogelijk nog een plot plaatsen van de normale verdeling

### cleaning the data for the moderate effect
The "imdb_movies.csv" dataset will be used for the moderate effect as well. The data cleaning of the moderate effect is very similar to the data cleaning of the direct effect. For example, the 5% extreme values will similarly be ignored as in the direct effect to fufil the normality assumption. Na's will also be similarly removed and start_year is filtered from 1990 onwards. The moderate dataset makes sure that every genre has it own row. This is managed through the seperate row function that is used. Genres are also filtered by "Drama" and "Action" because these are the genres that we focus on in our research. Finally, four columns are added in the moderate effect dataset: total_runtime, avg_rating, n_films and most_common_genre. The first three columns are created in the same way as in the direct effect dataset. The most_common_genre column and it provides the most common genres. All adjustments will be saved as csv file called "imdb_movies_mod.csv". 


# WE CREATED ANALYSIS AND VISUALIZATIONS BY USING THE FOLLOWING PACKAGES: DPLYR, GGALLY, GGPLOT2, TIDYR

```{r analysis_packagesecho = FALSE, message=FALSE, warning=FALSE}
#Loading packages
library(dplyr)
library(GGally)
library(ggplot2)
library(tidyr)
library()
```

#WE CREATED SOME MULTIPLE LINEAR REGRESSIONS, WHICH YOU CAN SEE IN THIS ANALYSIS

```{r regression_analysis}
# input from the dataprep_directeffect and dataprep_modeffect
readr::write_csv(director_stats, "directeffect_data.csv")
readr::write_csv(data_mod, "modeffect_data.csv")

# Creating a new variable for director_stats
director_stats$total_runtime <- director_stats$avg_runtime * director_stats$film_count

# Regressionmodel
model <- lm(avg_rating ~ total_runtime, data = director_stats)
summary(model)



# Regression with the moderator: first regression model integration with the moderator
data_mod$dominant_genre <- factor(data_mod$dominant_genre, levels = c("Drama", "Action"))

lrmmod <- lm(avg_rating ~ total_runtime * dominant_genre, data = data_mod)
summary(lrmmod)

#Output results from the lrmmod
lrmmodel_summary <- summary(lrmmod)
coef_lrmmod <- as.data.frame(lrmmod_summary$coefficients)
coef_lrmmod$Term <- ownames(coef_lrmmod)
coef_lrmmod <- coef_lrmmode\[, c("Term", "Estimate", "Std. Error", "t value", "Pr(>|t|)")\]
write.csv(coef_lrmmod,"analysis.csv", row.names = FALSE)
```
#CREATE THE CONCLUSION WE CAN DRAW FROM THIS ANALYSIS

#BELOW WE CREATED SOME VISUALIZATIONS TO LET YOU SEE OUR REGRESSION MODEL

```{r visualization_analysis}
# ggpairs plot
ggpairs(director_stats,
        columns = c("avg_rating", "film_count", "avg_runtime"),
        title = "Director Statistics")

# Creating a new variable for director_stats
director_stats$total_runtime <- director_stats$avg_runtime * director_stats$film_count

# Scatterplot with regressionline
ggplot(director_stats %>% dplyr::filter(total_runtime <= 20000),
       aes(x = total_runtime, y = avg_rating)) +
  geom_point(color = "blue", alpha = 0.6, size = 3) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    x = "Total runtime (avg_runtime * film_count)",
    y = "Average rating",
    title = "Linear Regression: Runtime vs Rating (≤ 20,000)"
  ) +
  theme_minimal()
```


#SHOW A CONCLUSION